---
title: "EDA"
format: pdf
editor: visual
---

```{r}
#| message: false
library(tidyverse)    
library(zoo)           
library(ggplot2)       
library(gridExtra)     
library(scales)       
library(lme4)         
library(geepack)      
library(viridis)  
library(here)
```

# input data

```{r}


data <- read.csv(here("Data/nursebp.csv"), header = TRUE)

data$SNUM <- as.factor(data$SNUM)
data$PHASE <- as.factor(data$PHASE)
data$DAY <- as.factor(data$DAY)
data$POSTURE <- as.factor(data$POSTURE)
data$FH123 <- as.factor(data$FH123)

data <- data %>%
  mutate(
    # in hours
    hour_of_day = floor(time / 60),
    #  20-minute 
    time_20 = floor(time / 20) * 20/60,
    time2 = time^2
  )
```

# Number of unique subjects

```{r}
n_subjects <- length(unique(data$SNUM))
n_subjects #有几个护士参与
```

# Average observations per subject

```{r}
nrow(data)/n_subjects
#每个人平均有几个观测值（给自己测了几次）
```

# Missing values by column

```{r}
missing_values <- sapply(data, function(x) sum(is.na(x)))

print(missing_values[missing_values > 0])
```

# Summary statistics

```{r}
num_vars <- c("SYS", "DIA", "HRT", "MNACT5", "STR", "HAP", "TIR", "AGE")
summary_stats <- data %>%
  select(all_of(num_vars)) %>%
  summary()
print(summary_stats)
```

```{r}
vars <- c("PHASE", "DAY", "POSTURE", "FH123")
for (var in vars) {
  cat(var)
  print(table(data[[var]]))
  cat("Percentages")
  print(round(prop.table(table(data[[var]])) * 100, 1))
}

```

#box plot

```{r}

par(mfrow = c(2, 2))
hist(data$SYS, main = "Distribution of Systolic BP", xlab = "SYS", col = "lightblue")
hist(data$DIA, main = "Distribution of Diastolic BP", xlab = "DIA", col = "lightgreen")
hist(data$HRT, main = "Distribution of Heart Rate", xlab = "HRT", col = "lightpink")
hist(data$MNACT5, main = "Distribution of Activity Level", xlab = "MNACT5", col = "lightyellow")
par(mfrow = c(1, 1))
```

```{r}
bp_by_day <- ggplot(data, aes(x = DAY, y = SYS, fill = DAY)) +
  geom_boxplot() +
  labs(title = "Systolic BP by Day Type", 
       x = "Day Type", 
       y = "Systolic Blood Pressure") +
  scale_fill_manual(values = c("skyblue", "salmon"),
                    labels = c("Non-workday", "Workday")) +
  theme_minimal()

print(bp_by_day)
```

```{r}

bp_by_fh <- ggplot(data, aes(x = FH123, y = SYS, fill = FH123)) +
  geom_boxplot() +
  labs(title = "Systolic BP by Family History", 
       x = "Family History", 
       y = "Systolic Blood Pressure") +
  scale_fill_manual(values = c("lightblue", "lightgreen", "salmon"),
                   labels = c("No hypertensive parents", 
                              "One hypertensive parent", 
                              "Both parents hypertensive")) +
  theme_minimal()



print(bp_by_fh)
```

```{r}

bp_by_str <- ggplot(data, aes(x = STR, y = SYS, fill = as.factor(STR))) +
  geom_boxplot() +
  labs(title = "Systolic BP by Family History", 
       x = "Family History", 
       y = "Systolic Blood Pressure") +
  scale_fill_manual(values = c("lightblue", "lightgreen", "salmon","blue","yellow"),
                   labels = c("1","2","3","4","5")) +
  theme_minimal()



print(bp_by_str)
```

# Correlation

```{r}


corr_vars <- c("SYS", "DIA", "HRT", "MNACT5", "STR", "HAP", "TIR", "AGE")
correlation_matrix <- cor(data[, corr_vars], use = "pairwise.complete.obs")
print(correlation_matrix[1,])

corrplot::corrplot(correlation_matrix, 
                   method = "color", 
                   type = "upper", 
                   order = "hclust",
                   tl.col = "black", 
                   tl.srt = 45,
                   addCoef.col = "black",
                   number.cex = 0.7)
```

#bars

```{r}

fh_summary <- data %>%
  group_by(FH123) %>%
  summarise(
    mean_SYS = mean(SYS, na.rm = TRUE),
    sd_SYS = sd(SYS, na.rm = TRUE),
    n = n(),
    se_SYS = sd_SYS / sqrt(n)
  )

print(fh_summary)
```

```{r}
subject_fh_data <- data %>%
  group_by(SNUM, FH123) %>%
  summarise(mean_SYS = mean(SYS, na.rm = TRUE)) %>%
  ungroup()

subject_fh_summary <- subject_fh_data %>%
  group_by(FH123) %>%
  summarise(
    mean_SYS = mean(mean_SYS, na.rm = TRUE),
    sd_SYS = sd(mean_SYS, na.rm = TRUE),#???? 为什么不行
    n = n(),
    se_SYS = sd_SYS / sqrt(n)
  )
print(subject_fh_summary)
```

```{r}

fh_plot <- ggplot(subject_fh_summary, aes(x = FH123, y = mean_SYS, fill = FH123)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7) +
  geom_errorbar(aes(ymin = mean_SYS, ymax = mean_SYS),
                width = 0.25, position = position_dodge(0.7)) +
  labs(title = "Mean Systolic BP by Family History (Subject-Level)",
       x = "Family History of Hypertension",
       y = "Mean Systolic Blood Pressure (mmHg)") +
  scale_x_discrete(labels = c("No history", "One parent", "Both parents")) +
  scale_fill_manual(values = c("lightblue", "lightgreen", "salmon"),
                    name = "Family History",
                    labels = c("No hypertensive parents", 
                               "One hypertensive parent", 
                               "Both parents hypertensive")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0))
print(fh_plot)
```

#Systolic BP by Stress Level

```{r}

stress_summary <- data %>%
  group_by(STR) %>%
  summarise(
    mean_SYS = mean(SYS, na.rm = TRUE),
    sd_SYS = sd(SYS, na.rm = TRUE),
    n = n(),
    se_SYS = sd_SYS / sqrt(n)
  )

print(stress_summary)
```

#stree level plot (no difference)

```{r}

stress_plot <- ggplot(stress_summary, aes(x = factor(STR), y = mean_SYS)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.7) +
  geom_errorbar(aes(ymin = mean_SYS - se_SYS, ymax = mean_SYS + se_SYS),
                width = 0.25) +
  labs(title = "Mean Systolic BP by Stress Level",
       x = "Stress Level",
       y = "Mean Systolic Blood Pressure (mmHg)") +
  theme_minimal()
print(stress_plot)
```

#Systolic BP by workday

```{r}

workday_summary <- data %>%
  group_by(DAY) %>%
  summarise(
    mean_SYS = mean(SYS, na.rm = TRUE),
    sd_SYS = sd(SYS, na.rm = TRUE),
    n = n(),
    se_SYS = sd_SYS / sqrt(n)
  )
print("Systolic BP by Working Status:")
print(workday_summary)
```

#small difference in workday

```{r}
workday_plot <- ggplot(workday_summary, aes(x = DAY, y = mean_SYS, fill = DAY)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7) +
  geom_errorbar(aes(ymin = mean_SYS - se_SYS, ymax = mean_SYS + se_SYS),
                width = 0.25, position = position_dodge(0.7)) +
  labs(title = "Mean Systolic BP by Working Status",
       x = "Day Type",
       y = "Mean Systolic Blood Pressure (mmHg)") +
  scale_x_discrete(labels = c("Non-workday", "Workday")) +
  scale_fill_manual(values = c("skyblue", "salmon"),
                    name = "Day Type",
                    labels = c("Non-workday", "Workday")) +
  theme_minimal()
print(workday_plot)
```

#Average change over the day

```{r}
hourly_summary <- data %>%
  group_by(hour_of_day) %>%
  summarise(
    mean_SYS = mean(SYS, na.rm = TRUE),
    sd_SYS = sd(SYS, na.rm = TRUE),
    n = n(),
    se_SYS = sd_SYS / sqrt(n)
  ) %>%
  filter(n >= 5) 
print(hourly_summary)
```

```{r}


time_plot <- ggplot(hourly_summary, aes(x = hour_of_day, y = mean_SYS)) +
  geom_line(size = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_ribbon(aes(ymin = mean_SYS - se_SYS, ymax = mean_SYS + se_SYS),
              alpha = 0.2, fill = "steelblue") +
  labs(title = "Diurnal Pattern of Systolic BP",
       x = "Hour of Day",
       y = "Mean Systolic Blood Pressure (mmHg)") +
  scale_x_continuous(breaks = seq(5, 23, 2)) +
  theme_minimal()
print(time_plot)


```

```{r}

time20_summary <- data %>%
  group_by(time_20) %>%
  summarise(
    mean_SYS = mean(SYS, na.rm = TRUE),
    sd_SYS = sd(SYS, na.rm = TRUE),
    n = n(),
    se_SYS = sd_SYS / sqrt(n)
  )


time20_plot <- ggplot(time20_summary, aes(x = time_20, y = mean_SYS)) +
  geom_line(size = 1, color = "darkblue") +
  geom_ribbon(aes(ymin = mean_SYS - se_SYS, ymax = mean_SYS + se_SYS),
              alpha = 0.2, fill = "steelblue") +
  labs(title = "Systolic BP Throughout the Day (20-minute intervals)",
       x = "Time (hours from midnight)",
       y = "Mean Systolic Blood Pressure (mmHg)") +
  theme_minimal()
print(time20_plot)


```

#workday ggplot

```{r}

time_workday <- data %>%
  group_by(hour_of_day, DAY) %>%
  summarise(
    mean_SYS = mean(SYS, na.rm = TRUE),
    sd_SYS = sd(SYS, na.rm = TRUE),
    n = n(),
    se_SYS = sd_SYS / sqrt(n)
  ) %>%
  filter(n >= 5) 


time_day_plot <- ggplot(time_workday, aes(x = hour_of_day, y = mean_SYS, color = DAY, group = DAY)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = mean_SYS - se_SYS, ymax = mean_SYS + se_SYS, fill = DAY),
              alpha = 0.2, color = NA) +
  labs(title = "Systolic BP by Time of Day and Working Status",
       x = "Hour of Day",
       y = "Mean Systolic Blood Pressure (mmHg)") +
  scale_x_continuous(breaks = seq(5, 23, 2)) +
  scale_color_manual(values = c("skyblue", "salmon"),
                    name = "Day Type",
                    labels = c("Non-workday", "Workday")) +
  scale_fill_manual(values = c("skyblue", "salmon"),
                    name = "Day Type",
                    labels = c("Non-workday", "Workday")) +
  theme_minimal()
print(time_day_plot)
```

#Family history ggplot

```{r}
time_fh_summary <- data %>%
  group_by(hour_of_day, FH123) %>%  
  summarise(
    mean_SYS = mean(SYS, na.rm = TRUE),
    sd_SYS = sd(SYS, na.rm = TRUE),
    n = n(),
    se_SYS = sd_SYS / sqrt(n),
    .groups = "drop"  
  ) %>%
  filter(n >= 5) 

time_fh_plot <- ggplot(time_fh_summary, aes(x = hour_of_day, y = mean_SYS, color = FH123, group = FH123)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = mean_SYS - se_SYS, ymax = mean_SYS + se_SYS, fill = FH123),
              alpha = 0.2, color = NA) +
  labs(title = "Systolic BP by Time of Day and Family History",
       x = "Hour of Day",
       y = "Mean Systolic Blood Pressure (mmHg)") +
  scale_x_continuous(breaks = seq(5, 23, 2)) +
  scale_color_manual(values = c("royalblue", "forestgreen", "firebrick"),
                    name = "Family History",
                    labels = c("No hypertensive parents", 
                               "One hypertensive parent", 
                               "Both parents hypertensive")) +
  scale_fill_manual(values = c("royalblue", "forestgreen", "firebrick"),
                    name = "Family History",
                    labels = c("No hypertensive parents", 
                               "One hypertensive parent", 
                               "Both parents hypertensive")) +
  theme_minimal()

print(time_fh_plot)
```

# Between-Subject vs. Within-Subject Variation

```{r}

subject_means <- data %>%
  group_by(SNUM) %>%
  summarise(mean_SYS = mean(SYS, na.rm = TRUE))

overall_mean <- mean(subject_means$mean_SYS)
between_subject_sd <- sd(subject_means$mean_SYS)

#Between-subject variation
#Overall mean SYS
overall_mean
#SD of subject means 
between_subject_sd

subject_hist <- ggplot(subject_means, aes(x = mean_SYS)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  geom_vline(xintercept = overall_mean, color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Subject Mean Systolic BP",
       x = "Mean Systolic BP (mmHg)",
       y = "Number of Subjects") +
  annotate("text", x = overall_mean + 5, y = 20, 
           label = paste("Overall Mean =", round(overall_mean, 1), "mmHg"),
           color = "red") +
  theme_minimal()
print(subject_hist)
```

```{r}


me_model <- lmer(SYS ~ time + DAY + HRT + MNACT5 + POSTURE + FH123 + TIR + (1 | SNUM), 
                data = data)
print(summary(me_model))

me_model2 <- lmer(SYS ~ time + time2 + DAY + HRT + MNACT5 + POSTURE + FH123 + TIR + (1 | SNUM), 
                 data = data)
print(summary(me_model2))
```

# table

#summary table

| Variable | Min | Max   | Mean  | Median |
|----------|-----|-------|-------|--------|
| SYS      | 75  | 200   | 118.2 | 117    |
| DIA      | 40  | 120   | 71.4  | 71     |
| HRT      | 35  | 144   | 80.0  | 80     |
| MNACT5   | 0   | 359.4 | 190.4 | 207    |
| STR      | 1   | 5     | 1.5   | 1      |
| HAP      | 1   | 5     | 3.1   | 3      |
| TIR      | 1   | 5     | 2.0   | 2      |
| AGE      | 24  | 50    | 37.8  | 38     |

# distribution table

| Variable | Category         | Count | Percentage |
|----------|------------------|-------|------------|
| PHASE    | L (luteal)       | 4,836 | 50.5%      |
|          | F (follicular)   | 4,737 | 49.5%      |
| DAY      | W (workday)      | 5,457 | 57.0%      |
|          | NW (non-workday) | 4,116 | 43.0%      |
| POSTURE  | SIT              | 4,101 | 45.6%      |
|          | STAND            | 4,255 | 47.3%      |
|          | RECLINE          | 631   | 7.0%       |
| FH123    | NO               | 5,298 | 55.3%      |
|          | YES              | 3,633 | 38.0%      |
|          | YESYES           | 642   | 6.7%       |

#Family history table by nurse

| Family History                     | Mean SYS | Count | SE   |
|------------------------------------|----------|-------|------|
| NO (no hypertensive parents)       | 117.6    | 112   | 0.79 |
| YES (one hypertensive parent)      | 117.6    | 77    | 0.96 |
| YESYES (both parents hypertensive) | 125.7    | 14    | 2.40 |

\

\

\

# Model 1 vs Model 3 Fixed Effects Between Models

## Table 1: Comparison of Fixed Effects Between Models

+-----------------+-------------------+---------------+-------------------+---------------+
| Variable        | Model1            |               | Model2            |               |
|                 |                   |               |                   |               |
|                 | (with DIA)        |               | (without DIA)     |               |
+=================+===================+===============+===================+===============+
|                 | **Estimate (SE)** | **p-value**   | **Estimate (SE)** | **p-value**   |
+-----------------+-------------------+---------------+-------------------+---------------+
| Intercept       | 72.86 (4.21)      | \<0.001\*\*\* | 99.48 (4.87)      | \<0.001\*\*\* |
+-----------------+-------------------+---------------+-------------------+---------------+
| AGE             | -0.031 (0.074)    | 0.679         | 0.070 (0.092)     | 0.447         |
+-----------------+-------------------+---------------+-------------------+---------------+
| DIA             | 0.492 (0.012)     | \<0.001\*\*\* | —                 | —             |
+-----------------+-------------------+---------------+-------------------+---------------+
| FH123-YES       | -0.729 (1.015)    | 0.474         | -0.787 (1.265)    | 0.534         |
+-----------------+-------------------+---------------+-------------------+---------------+
| FH123-YESYES    | 4.548 (1.921)     | 0.019\*       | 7.108 (2.391)     | 0.003\*\*     |
+-----------------+-------------------+---------------+-------------------+---------------+
| HRT             | 0.040 (0.014)     | 0.006\*\*     | 0.099 (0.016)     | \<0.001\*\*\* |
+-----------------+-------------------+---------------+-------------------+---------------+
| MNACT5          | 0.032 (0.002)     | \<0.001\*\*\* | 0.033 (0.003)     | \<0.001\*\*\* |
+-----------------+-------------------+---------------+-------------------+---------------+
| STR2            | 0.217 (0.359)     | 0.606         | 0.512 (0.396)     | 0.196         |
+-----------------+-------------------+---------------+-------------------+---------------+
| STR3            | -0.543 (0.579)    | 0.938         | -0.293 (0.640)    | 0.646         |
+-----------------+-------------------+---------------+-------------------+---------------+
| STR4            | 1.224 (0.970)     | 0.262         | 2.044 (1.069)     | 0.056         |
+-----------------+-------------------+---------------+-------------------+---------------+
| STR5            | 0.591 (1.681)     | 0.725         | 3.004 (1.850)     | 0.105         |
+-----------------+-------------------+---------------+-------------------+---------------+
| TIR             | -0.059 (0.185)    | 0.749         | -0.319 (0.208)    | 0.126         |
+-----------------+-------------------+---------------+-------------------+---------------+
| DAYW            | 1.166 (0.960)     | 0.226         | 2.703 (1.194)     | 0.025\*       |
+-----------------+-------------------+---------------+-------------------+---------------+
| PHASEL          | 0.666 (0.959)     | 0.489         | 0.412 (1.195)     | 0.731         |
+-----------------+-------------------+---------------+-------------------+---------------+
| POSTURE-RECLINE | 0.574 (2.731)     | 0.833         | -3.629 (2.999)    | 0.226         |
+-----------------+-------------------+---------------+-------------------+---------------+
| POSTURE-SIT     | 1.158 (2.670)     | 0.665         | 0.133 (2.932)     | 0.964         |
+-----------------+-------------------+---------------+-------------------+---------------+
| POSTURE-STAND   | 1.186 (2.669)     | 0.657         | 0.512 (2.931)     | 0.861         |
+-----------------+-------------------+---------------+-------------------+---------------+
| time            | 0.192 (0.200)     | 0.338         | 0.063 (0.244)     | 0.796         |
+-----------------+-------------------+---------------+-------------------+---------------+
| $\text{time}^2$ | 0.137 (0.152)     | 0.367         | 0.354 (0.168)     | 0.035\*       |
+-----------------+-------------------+---------------+-------------------+---------------+

\

\

\

\
\# Randome effect model 1 vs model 3 Random Effects Components

## Table 2: Random Effects Components

+---------------+--------------+------------+---------------+------------+
| Random Effect | Model1       |            | Model2        |            |
|               |              |            |               |            |
|               | (with DIA)   |            | (without DIA) |            |
+===============+==============+============+===============+============+
|               | **Variance** | **SD**     | **Variance**  | **SD**     |
+---------------+--------------+------------+---------------+------------+
| Intercept     | 38.78        | 6.23       | 61.04         | 7.81       |
+---------------+--------------+------------+---------------+------------+
| time slope    | 3.20         | 1.79       | 5.80          | 2.41       |
+---------------+--------------+------------+---------------+------------+
| Residual      | 126.04       | 11.23      | 151.32        | 12.30      |
+---------------+--------------+------------+---------------+------------+
| Correlation   | 0.25         | —          | 0.21          | —          |
+---------------+--------------+------------+---------------+------------+
