---
title: "Main Analysis"
author: "Group C"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Description
The UCLA Nurse Blood Pressure Study collected information from registered nurses in the Los Angeles area between 24 and 50 years of age on blood pressure (BP) and potential factors that contribute to hypertension. This information includes family history, and whether the subject had one or two hypertensive parents, as well as a wide range of measures of the physical and emotional condition of each nurse throughout the day. Researchers sought to study the links between BP and family history, personality, mood changes, working status, and menstrual phase. The first BP measurement was taken half an hour before the subject’s normal start of work, and BP was then measured approximately every 20 minutes for the rest of the day. At each BP reading, the nurses also rate their mood on several dimensions, including how stressed they feel at the moment the BP is taken. In addition, the activity of each subject during the 10 minutes before each reading was measured using an
actigraph worn on the waist.

## Research Question
What are the factors that are associated with higher systolic blood pressure?

## Load necessary packages
```{r, message=FALSE}
library(here)
library(tidyverse)
library(dplyr)
library(car)
library(mice)
library(lme4)
library(nlme)
library(lmerTest)
library(ggplot2)
```

## Data Inspection and Manipulation 
```{r}
# Load data
nursebp <- read.csv(here("Data/nursebp.csv"), header = TRUE)

# Check structure
str(nursebp)

# Summary of key variables
summary(nursebp)

# Count missingness
colSums(is.na(nursebp))

# Number of unique subjects
length(unique(nursebp$SNUM))

nursebp <- nursebp %>%
  mutate(
    DAY = as.factor(DAY),
    FH123 = as.factor(FH123),
    PHASE = as.factor(PHASE),
    POSTURE = as.factor(POSTURE),
    SNUM = as.factor(SNUM),
    STR = as.factor(STR),
    HAP = as.factor(HAP)
  )
```

## Exploratory Analysis
```{r}
ggplot(nursebp, aes(x = SYS)) +
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.6) +
  labs(title = "Distribution of Systolic Blood Pressure",
       x = "SYS (mmHg)", y = "Count")

ggplot(nursebp, aes(x = time, y = SYS, group = SNUM, color = SNUM)) +
  geom_line(alpha = 0.4) +
  labs(title = "Individual SBP Trajectory Over Time",
       x = "Time (minutes from midnight)", y = "SYS (mmHg)") +
  theme(legend.position = "none")

ggplot(nursebp, aes(x = FH123, y = SYS, fill = FH123)) +
  geom_boxplot() +
  labs(title = "SBP by Family History of Hypertension",
       x = "Family History", y = "SYS (mmHg)")

ggplot(nursebp, aes(x = PHASE, y = SYS, fill = PHASE)) +
  geom_boxplot() +
  labs(title = "SBP by Menstrual Phase",
       x = "Menstrual Phase", y = "SYS (mmHg)")

ggplot(nursebp, aes(x = POSTURE, y = SYS, fill = POSTURE)) +
  geom_boxplot() +
  labs(title = "SBP by Posture",
       x = "Position during BP measurement", y = "SYS (mmHg)")
```

Correlation among variables
```{r}
# DIA should be correlated with SYS, and it is with moderate correlation. The other variables seem no obvious correlation.
nursebp %>% 
  select(AGE, SYS, DIA, HRT, time) %>% 
  cor(use = "complete.obs")
```

## Modelling
Missingness examination
```{r}
nursebp$STR_missing <- ifelse(is.na(nursebp$STR), 1, 0)
table(nursebp$AGE, nursebp$STR_missing)

# Density plot of age by missingness indicator
ggplot(nursebp, aes(x = AGE, fill = factor(STR_missing))) +
  geom_density(alpha = 0.5) +
  labs(x = "Age",
       y = "Density",
       fill = "STR Missing (1 = Yes, 0 = No)",
       title = "Age Distribution by STR Missingness")

## still need to decide whether impute the data
```

```{r}
fixed_model <- lm(SYS ~ AGE + DIA + FH123 + HAP + HRT + MNACT5 + STR + TIR + DAY + PHASE + POSTURE + time + I(time^2), data = nursebp)
vif(fixed_model)
## high multicollinearity among time and time^2, could consider centering
```

1. Include all variables
```{r}
model0 <- lmer(SYS ~ AGE + DIA + FH123 + HAP + HRT + MNACT5 + STR + TIR + DAY + PHASE + POSTURE + time + (1 + time | SNUM), data = nursebp)
```
2. Remove DIA
```{r}
model1 <- lmer(SYS ~ AGE + FH123 + HAP + HRT + MNACT5 + STR + TIR + DAY + PHASE + POSTURE + time + (1 + time | SNUM), data = nursebp)
anova(model0, model1)
# Based on likelihood ratio test, we should not remove DIA
# Just a test for model comparison
```

```{r}
# Center and scale time
nursebp$time_c <- scale(nursebp$timepass, center = TRUE, scale = TRUE)
model0 <- lmer(SYS ~ AGE + DIA + FH123 + HAP + HRT + MNACT5 + STR + TIR + DAY + PHASE + POSTURE + time_c + (1+ time_c|SNUM), data = nursebp)
summary(model0)
```

```{r}
model1 <- lmer(SYS ~ AGE + DIA + FH123 + HAP + HRT + MNACT5 + STR + TIR + DAY + PHASE + POSTURE + time_c + (1|SNUM), data = nursebp)
summary(model1)
```

```{r}
#Random intercept and covariance pattern model
nursebp_lme <- na.omit(nursebp)

#AR1 error
nursebp_lme$time_c <- as.integer(rank(nursebp_lme$time_c))
fit1 <- lme(SYS ~ AGE + DIA + FH123 + HAP + HRT + MNACT5 + STR + TIR + DAY + PHASE + POSTURE + time_c,random = ~time_c | SNUM,
            correlation = corAR1(form = ~time_c| SNUM),
            data = nursebp_lme)
```

```{r}
#exponential error
fit2 <- lme(SYS ~ AGE + DIA + FH123 + HAP + HRT + MNACT5 + STR + TIR + DAY + PHASE + POSTURE + time_c,random = ~time_c | SNUM,
            corr=corExp(form = ~time_c|SNUM),
            data = nursebp_lme)
```

```{r}
#heteroskedasticity error cross time
fit3 <- lme(SYS ~ AGE + DIA + FH123 + HAP + HRT + MNACT5 + STR + TIR + DAY + PHASE + POSTURE + time_c,random = ~time_c | SNUM,
            weights=varIdent(form = ~ 1|time_c), 
            data = nursebp_lme)
```

```{r}
#heteroskedasticity error cross genetic factor
fit4 <- lme(SYS ~ AGE + DIA + FH123 + HAP + HRT + MNACT5 + STR + TIR + DAY + PHASE + POSTURE + time_c,random = ~time_c | SNUM,
            weights=varIdent(form = ~ 1|FH123), 
            data = nursebp_lme)
```

```{r}
# Create a data frame to store the model comparison metrics
model_comparison <- data.frame(
  Model = c("AR1 error", "Exponential error",  "Heteroskedastic genetic"),
  AIC = c(AIC(fit1), AIC(fit2),  AIC(fit4)),
  BIC = c(BIC(fit1), BIC(fit2),  BIC(fit4))
)

# If you want to include log-likelihood also
model_comparison$logLik <- c(logLik(fit1)[1], logLik(fit2)[1], logLik(fit4)[1])

# Sort models by AIC (lowest is best)
model_comparison <- model_comparison[order(model_comparison$AIC), ]

# Print the formatted table
print(model_comparison, digits = 2)

# If you want a nicer table for reports/documents
library(knitr)
kable(model_comparison, digits = 2, caption = "Model Comparison by Information Criteria")
```

## Diagnostics
```{r}

```

## Results and Interpretation
```{r}

```

## Visualization
```{r}

```
